---
title: "SumStat"
author: "gc5k"
date: "7/7/2018"
output: 
  html_document:
    theme: united
    highlight: tango
    code_folding: hide
---


```{r, date, echo=FALSE, warning=FALSE}
Date=Sys.Date()[1]
print(paste0("Document was last updated at ", Date, "."))
```

----

#Table of contents {.tabset .tabset-fade .tabset-pills}

##1 Theory


**Background**

Several studies have used summary statistics for the inference of genetic architecture 

1. [AJHG, 2014, 96:21-36](Am J Hum Genet, 2014, 96:21-36), 

2. [PLoS Genet, 2015,10:e1004198](http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3967938&tool=pmcentrez&rendertype=abstract)

3. [Hum Mol Genet, 2016, 25:1857-1866](https://academic.oup.com/hmg/article/25/9/1857/2385201?searchresult=1)

4. [Am Stat, 2017](https://www.tandfonline.com/doi/full/10.1080/00031305.2016.1277159)

5. [BMC Genomics, 2018, 19:494](https://www.ncbi.nlm.nih.gov/pubmed/29940862)

--------------------------------------

**Theory**

The theory of the paper can be found in [*Eur J Hum Genet*, 2017, 25:137-46](https://www.nature.com/articles/ejhg2016106). For a pair of cohort their correlation of their z scores can be defined 

$\rho=cor(\textbf{z}_1,\textbf{z}_2)=\frac{\frac{\sqrt{h^2_1h^2_2}}{m}\gamma_G+\frac{\gamma_{o}}{\sqrt{n_1n_2}}\gamma_P}{\sqrt{\frac{h^2_1}{m}+\frac{1}{n_1}}\sqrt{\frac{h^2_2}{m}+\frac{1}{n_2}}}$

|Notation|Interpretation|
|--------|--------------------------------------------|
|$\textbf{z}$| z scores, a vector of $M\times1$, for SNPs|
|$h^2_{.}$|heritbility for a trait|
|$m$|SNPs|
|$n_{.}$|Sample size|
|$\gamma_G$|Genetic correlation for a pair of traits|
|$\gamma_P$|Phenotypic correlation for a pair of traits|
|$r_{o}=\frac{n_o}{\sqrt{n_1n_2}}$|Overlapping samples between a pair of cohorts|

##2 Simulating a cohort

**Parameter setting (no LD):**

1. Sample size, $n=\{1000, 2000\}$
2. number of markers, $m = \{1,000, 10,000\}$
3. number of QTLs, $m_{Q} = \{100, 500, 1,000\} (<=m)$
4. For $m$ loci, its allele frequency is from uniform distribution $p \sim U(0.05, 0.95)$
5. Genetic effect is sampled from $\mathbf{\beta}_i \sim N(0, \frac{h^2}{m_Q})$
6. heritability, $h^2=\{0.1, 0.2, 0.5\}$

The whole genotype matrix is $\mathbf{G}_{n\times m}$. For individual $i$ its $j^{th}$ locus $\mathbf{G}_{ij}$ is sampled from binomial distribution $\mathbf{B}(p_j,2)$, in which $p_j$ is the allele frequency of the locus.

$\mathbf{y}=\mu+\mathbf{G}\beta+e$

in which $e \sim N(0, \frac{\sigma^2_{(\mathbf{G\beta})}}{h^2}(1-h^2))$ so that the genotypic variation $V_G=Var(\mathbf{G}\beta)$ can be scaled to satify heritability.

```{r, generating a single cohort}
print("Write R code for simulating a single cohort")
```

------------------------------

**Estimate genetic effects**

$\mathbf{y}=a+\beta_j\mathbf{g}_j+e$

```{r, gwas}
print("Write R code for estimating single-marker gwas")
```

After GWAS, we have a table for GWAS summary statistics

| SNP | Ref Allele | Alt Allele | Freq | Effect | SE |
|---|----|-----|----|----|----|
|$SNP_1$ | G | A | $\hat{p}_1$ | $\hat\beta_1$ | $\hat{\sigma}_{\beta_1}$ |
|$SNP_2$ | T | C | $\hat{p}_2$ | $\hat\beta_2$ | $\hat{\sigma}_{\beta_2}$ |
|$\mathbf{SNP}_3$ | ~~T~~ | ~~A~~ | $\hat{p}_3$ | $\hat\beta_3$ | $\hat{\sigma}_{\beta_3}$ |
|$\mathbf{SNP}_4$ | ~~G~~ | ~~C~~ | $\hat{p}_4$ | $\hat\beta_4$ | $\hat{\sigma}_{\beta_4}$ |
|$\vdots$| $\vdots$ | $\vdots$| $\vdots$ | $\vdots$ | $\vdots$ |
|$SNP_m$ | C | T|  $\hat{p}_m$ | $\hat\beta_m$ | $\hat{\sigma}_{\beta_m}$ |
|||||||


##3 Simulating a pair of cohorts

In general, the procedure is the same as for simulating a single cohort. 

However, for cohort 2 that is derived from the same ancestry as cohort 1, the expected allele frequency $p_{j,2}=p_{j,1}$. 

Given the genetic correlation between cohort 1 and 2 is $cor(\mathbf{\beta}_{1}, \mathbf{\beta}_{2})=\gamma_G$, the genetic effect for cohort 2 can be simulated as $\beta_2=\gamma_G\beta_1+\sqrt{1-r^2}\tilde{e}$, in which $\tilde{e} \sim (0, h^2_2)$.

--------------------

**Directly simulating summary statistics for GWAS**

Assuming the genetic effect is estimated via LSE for the model below

$\mathbf{y}=a+\beta_i \mathbf{g}_i+e$ 

For locus $j$, its estimated effect is distributed
$\hat\beta_i \sim N(\beta_i, \sqrt{\frac{\sigma_y^2-\sigma_{g_i}^2 \beta^2_i}{(n_1-n_u)\sigma_{g_i}^2}})$, in which $E(\sigma_{g_i}^2)=2p_i \bar p_i$, and $\hat{\sigma}_{\beta}^2=\frac{\sigma_y^2-\sigma_{g_i}^2 \beta^2_i}{(n_1-n_u)\sigma_{g_i}^2}$, $n_u$ ($n_c=2$ here for the model above) is the number of parameters in the LSE.

For a pair of cohorts we have summary statistics below

| SNP | Ref Allele$_1$ | Alt Allele$_1$ | Freq$_1$ | Effect$_1$ | SE$_1$ | Ref Allele$_2$ | Alt Allele$_2$ | Effect$_2$ | Freq$_2$ | SE$_2$ |
|:---|:------------:|:------------:|:---------------:|:----------:|:----------:|:----------------:|:---------------:|:-----------:|:-------:|:-------:|
|$SNP_{1}$ | A | C | $\hat{p}_{1.1}$ | $\hat\beta_{1.1}$ | $\hat{\sigma}_{\beta_{1.1}}$ |  A | C | $\hat{p}_{1.1}$ | $\hat\beta_{1.2}$ | $\hat{\sigma}_{\beta_{1.2}}$ |
|$\mathbf{SNP}_{2}$ | ~~G~~ | ~~C~~ | $\hat{p}_{2.1}$ | $\hat\beta_{2.1}$ | $\hat{\sigma}_{\beta_{2.1}}$ |  ~~G~~ | ~~C~~ | $\hat{p}_{2.1}$ | $\hat\beta_{2.2}$ | $\hat{\sigma}_{\beta_{2.2}}$ |
|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|
|$SNP_{m}$ | A | C | $\hat{p}_{m.1}$ | $\hat\beta_{m.1}$ | $\hat{\sigma}_{\beta_{m.1}}$ |  A | C | $\hat{p}_{m.1}$ | $\hat\beta_{m.2}$ | $\hat{\sigma}_{\beta_{m.2}}$ |
||||||||||||


##4 Simulating nuclear families


##5 $\lambda_{meta}$

Given the summary statistics below, we have

| SNP | Ref Allele$_1$ | Alt Allele$_1$ | Freq$_1$ | Effect$_1$ | SE$_1$ | Ref Allele$_2$ | Alt Allele$_2$ | Effect$_2$ | Freq$_2$ | SE$_2$ |
|:---|:------------:|:------------:|:---------------:|:----------:|:----------:|:----------------:|:---------------:|:-----------:|:-------:|:-------:|
|$SNP_{1}$ | A | C | $\hat{p}_{1.1}$ | $\hat\beta_{1.1}$ | $\hat{\sigma}_{\beta_{1.1}}$ |  A | C | $\hat{p}_{1.1}$ | $\hat\beta_{1.2}$ | $\hat{\sigma}_{\beta_{1.2}}$ |
|$\mathbf{SNP}_{2}$ | ~~G~~ | ~~C~~ | $\hat{p}_{2.1}$ | $\hat\beta_{2.1}$ | $\hat{\sigma}_{\beta_{2.1}}$ |  ~~G~~ | ~~C~~ | $\hat{p}_{2.1}$ | $\hat\beta_{2.2}$ | $\hat{\sigma}_{\beta_{2.2}}$ |
|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|$\vdots$|
|$SNP_{m}$ | A | C | $\hat{p}_{m.1}$ | $\hat\beta_{m.1}$ | $\hat{\sigma}_{\beta_{m.1}}$ |  A | C | $\hat{p}_{m.1}$ | $\hat\beta_{m.2}$ | $\hat{\sigma}_{\beta_{m.2}}$ |
||||||||||||

We can estimate $\lambda_{meta}$ statistic as below

$h_i= \frac{(\beta_{j,1}-\beta_{j,2})^2}
{\sigma_{j,1}^2+\sigma_{j,2}^2} \sim \chi^2_1$

For $m$ loci, we have $\mathbf{h}$, a $m\times 1$ vector. As the vector $\mathbf{h} \sim \chi^2_1$, we can define $\lambda_{meta}=\frac{median(\mathbf{h})}{0.455}$, in which 0.455 is the quantity for $\chi^2_{1,p=0.5}=0.455$. Under the null hypothesis of no overlapping samples, $E(\lambda_{meta})=1$.

$\hat{\rho}_{\lambda_{meta}}=1-\lambda_{meta}$, and $E(\rho_{\lambda_{meta}})=\frac{n_o}{\sqrt{n_1n_2}}$.

##6 Various overlapping relatives

1. Overlapping individuals

2. Overlapping relatives

##7 Meta-analysis

For a pair of cohorts, for the$j^{th}$ locus, we have $\hat\beta_{j,1}$, $\hat{\sigma}^2_{\beta_{j,1}}$, and $\hat\beta_{j,1}$, $\hat{\sigma}^2_{\beta_{j,1}}$.

Using generalized (or weighted) least squares (GLS) estimates the genetic effect

$\mathbf{y}=\mu+e$

in which $\mathbf{y}^T=[\hat{\beta}_{j,1}, \hat{\beta}_{j,2}]$, and 
$$
e=
\left[
\begin{matrix} 
1&\rho_{1,2} \\
\rho_{1,2}&1
\end{matrix}
\right]
$$

Correct and incorrect ways in estimating 

Correct but subject to heterogeneity: $\hat\rho=1-\lambda_{meta}$

Incorrect and subject to genetic architecture: $\hat\rho=cov(\mathbf{Z}_1, \mathbf{Z}_2)$ or $\hat\rho=cov(\mathbf{\beta}_1, \mathbf{\beta}_2)$

More details for the estimation method can be [1 & 2 are generalized lse, 3 & 4 are chisq test]

1. Am J Hum Genet, "Meta-analysis of genome-wide association studies with overlapping subjects.", 2009:85, 862-72, **Eq 3-5**

2. Bioinformatics, "METAL: fast and efficient meta-analysis of genomewide association scans", 2010:26, 2190-1, See their **Table**

3. [Zhu Xiaofeng, Am J Hum Genet, 2014, 96:21-36](Am J Hum Genet, 2014, 96:21-36)

4. [Mike Goddard, PLoS Genet, 2015,10:e1004198](http://www.pubmedcentral.nih.gov/articlerender.fcgi?artid=3967938&tool=pmcentrez&rendertype=abstract)

5. [Wang et al, Hum Mol Genet, 2013, 22:2303-2311](https://academic.oup.com/hmg/article/22/11/2303/634133)

```{r, meta}
set.seed(1000)
CT=4
N=ceiling(rnorm(CT, 3000, 1000))
M=10000

#freq
P=runif(M, 0.05, 0.95)
Freq=matrix(0, M, CT)
for(i in 1:CT) {
  Freq[,i] = rnorm(M, P, sd=sqrt((P*(1-P))/(2*N[i])))
}

#mat
h2=0.5
Beta=rnorm(M, 0, sd=0)#=sqrt(h2/sum(2*P*(1-P))))
Bmat=matrix(0, M, CT)
Smat=matrix(0, M, CT)
Pmat=matrix(0, M, CT)
for(i in 1:CT) {
  Smat[,i]=1/(N[i]*2*Freq[,i]*(1-Freq[,i]))
  Bmat[,i] = rnorm(M, Beta, sd=sqrt(Smat[,i]))
  Pmat[,i] = (1-pnorm(abs(Bmat[,i]/sqrt(Smat[,i]))))*2
}

#fixed model
FisherP=matrix(0, M, 1)
Fmeta=matrix(0, M, 3)
Qmeta=matrix(0, M, 6)
for(i in 1:M) {
  FisherP[i,1]=-2*sum(log(Pmat[i,]))

  v=1/Smat[i,]
  Fmeta[i,2]=1/sum(v)
  Fmeta[i,1]=sum(Bmat[i,]*v)/sum(v)
  Fmeta[i,3]=Fmeta[i,1]/sqrt(Fmeta[i,2])

  Qmeta[i,4]=sum(v*(Bmat[i,]-Fmeta[i,1])^2)
  Qmeta[i,5]=sum(v)-sum(v^2)/sum(v)
  Qmeta[i,6]=(Qmeta[i,4]-(CT-1))/Qmeta[i,5]
  if(Qmeta[i,6]<0) {
    Qmeta[i,6]=0
  }
  vr=1/(Smat[i,]+Qmeta[i,6])
  Qmeta[i,2]=1/sum(v)+Qmeta[i,6]
  Qmeta[i,1]=sum(vr*Bmat[i,])/sum(vr)
  Qmeta[i,3]=Qmeta[i,1]^2/Qmeta[i,2]
}
layout(matrix(1:4, 2, 2))
qqplot(main="Fisher's methods", xlab=expression(paste("Theoretical ", chi[8]^2)),
       ylab=expression(paste("Observed ", chi[8]^2)),
       FisherP, rchisq(M, 8), bty="n", pch=16)
abline(a=0, b=1)

qqplot(main="Fixed model", xlab=expression(paste("Theoretical ", chi[1]^2)),
       ylab=expression(paste("Observed ", chi[1]^2)),
       rchisq(M, 1), Fmeta[,3]^2, bty='n', pch=16)
abline(a=0, b=1)

qqplot(main="Random model", xlab=expression(paste("Theoretical ", chi[1]^2)),
       ylab=expression(paste("Observed ", chi[1]^2)),
       rchisq(M,1), Qmeta[,1]^2/Qmeta[,2], bty='n', pch=16)
abline(a=0, b=1)

qqplot(main="Heterogeneity (4 cohorts)", xlab="Q", ylab=expression(paste("Theoretical ", chi[3]^2)), Qmeta[,4], rchisq(M,CT-1), bty='n', pch=16)
abline(a=0, b=1)
```

##8 MLM GWAMA
Mixed model for GWAMA

$\beta=\mu+Mb+e$

$\beta\sim N(\mu, \sigma_M+\sigma_e)$

##9 Notes for LD simulation

**A pair of loci**

|Notations|Interpretation|
|----|------------------------------------------------------------------|
|$p_A, p_a=\bar p_A$ | Allele frequency for $A$ and $a$ at locus **A**|
|$p_B, p_b=\bar p_B$ | Allele frequency for $B$ and $b$ at locus **B**|
|$p_{AB}$ | Haplotype frequency for $AB$ |
|$D_{AB}$ | Linkage disequilibrium $D_{AB}=p_{AB}-p_Ap_B$; the maximal $D_{AB}=0.25$ is reached when $p_A=p_B=0.5$|
|${D}_{AB}^\prime$|Standardized $D_{AB}$; if $D_{AB} > 0$, ${D}_{AB}^\prime=\frac{{D}_{AB}}{min(p_ap_B,p_Ap_b)}$, if $D_{AB} < 0$, ${D}_{AB}^\prime=\frac{{D}_{AB}}{min(p_ap_b,p_Ap_B)}$|
||

**Haplotype table**

| |  | Locus $\textbf{A}$ | |
|:---:|:----:|:-----:|:-----:|:--:|
|  |   | $A$  | $a$| |
|Locus $\textbf{B}$ | $B$ |$AB: p_Ap_B+D$ | $aB: p_ap_B-D$| $p_B$|
| |$b$|$Ab: p_Ap_b-D$  | $ab: p_ap_b+D$| $p_b$|
| | |$p_A$ | $p_a$ | |
||||||

In haplotype simulation, under random mating, the conditional probability of generating the second locus could be expressed as (t is generation)

1. $p(\mathbf{B}=B | \mathbf{A}=A)=\frac{p_Ap_B+D_{AB}^t}{p_A}$
2. $p(\mathbf{B}=b | \mathbf{A}=A)=\frac{p_Ap_b-D_{AB}^t}{p_A}$
3. $p(\mathbf{B}=B|\mathbf{A}=a)=\frac{p_ap_B-D_{AB}^t}{p_a}$
4. $p(\mathbf{B}=B|\mathbf{A}=a)=\frac{p_ap_b+D_{AB}^t}{p_a}$

$D_{AB}^t=D_{AB}(1-c_{AB})^t$, in which $c_{AB}$ is the recombination between loci **A** and **B**.

```{r, ld simulation}
print("Write R code for ld simulation")
```

**Reference**

Devlin B., Risch N., 1995 A comparison of linkage disequilibrium measures for fine-scale mapping. **Genomics** 29: 311-22.

Lewontin R. C., 1964 The Interaction of Selection and Linkage. I. General Considerations; Heterotic Models. **Genetics** 49: 49-67.
